Numerical Stability of Detonations in White Dwarf Simulations

### 2018-07-17 papers/ignition/paper.tex
Section -> subsection
Put the AMR discussion before the timestep limiter discussion

\subsection{Timestep Limiting}
\label{sec:timestep}

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
Consider the basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
The principal drawback of this approach is that the reactions and the hydrodynamics
can become decoupled from each other. For example, consider the example of a head-on
collision of white dwarfs. Initial nuclear burning will occur at the contact point
between the two stars. Because of the staggered updates from splitting, the evolution
effectively progresses as a cycle between burning for $\Delta t$ and getting fresh
material advected into the contact point by the hydro update for $\Delta t$.
When the collision begins, $\Delta t$ is controlled by the hydrodynamic stability
criterion, and may be large enough that it is possible for the burning advance in
Step 4 to completely burn the freshly advected material all the way to NSE.
Consequently the evolution is no longer a good approximation to smooth burning of
the in-falling material but rather separate discrete burning and hydro steps, and
the nature of the burning evolution will be quite different.

A common solution to this problem presented in the literature has been to limit the
size of the timestep and thereby limit the extent of this decoupling
\citep{raskin:2010,hawley:2012}, which we adopt here and have implemented in \castro.
Defining the nuclear energy injection timescale $\tau_e$, and the species evolution
timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations.
This limiter is complementary to the limiter based on $f_{be}$, and can catch
certain burning regimes that limiting only based on changes in energy would miss.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. We choose to do this with a direct evaluation of the
right-hand-side of the burning network, given the current state at the time when
the timestep is evaluated. Other work in the literature has typically evaluated
this as a finite difference using the change over the last timestep. Our method
is more accurate in principle, but in practice for this problem we find no
meaningful difference between the limiting we propose and other methods of
estimating the time derivatives.

The size of the splitting error is a function of the advance timestep, and it
decreases as the timestep becomes smaller. When we add AMR in the regions of
nuclear burning, the hydrodynamic timestep automatically becomes smaller due
to the increased resolution, which mitigates the splitting error. In fact, we
have found that when we have sufficient refinement to achieve a converged
simulation, the size of the hydrodynamic timestep is already small enough
that the splitting error is minimal. Consequently this timestep limiter
is only of great importance for low resolution simulations. Additionally,
we found that for the test problem described later, where we are only trying
to locate the point and time of a detonation rather than assess accurate
nucleosynthetic yields, this timestep limiter has no observable effect,
so we do not discuss it further in this paper.

\section{Timestep Retries}
\label{sec:retries}

>>>>

\subsection{Timestep Size}
\label{sec:timestep}

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
Consider the basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
The principal drawback of this approach is that the reactions and the hydrodynamics
can become decoupled from each other. For example, consider the example of a head-on
collision of white dwarfs. Initial nuclear burning will occur at the contact point
between the two stars. Because of the staggered updates from splitting, the evolution
effectively progresses as a cycle between burning for $\Delta t$ and getting fresh
material advected into the contact point by the hydro update for $\Delta t$.
When the collision begins, $\Delta t$ is controlled by the hydrodynamic stability
criterion, and may be large enough that it is possible for the burning advance in
Step 4 to completely burn the freshly advected material all the way to NSE.
Consequently the evolution is no longer a good approximation to smooth burning of
the in-falling material but rather separate discrete burning and hydro steps, and
the nature of the burning evolution will be quite different. This can have a
substantial effect on nucleosynthetic yields predicted by white dwarf simulations.

The size of the splitting error is a function of the advance timestep, and it
decreases as the timestep becomes smaller. When we add AMR in the regions of
nuclear burning, the hydrodynamic timestep (as set by the CFL criterion)
automatically becomes smaller due to the increased resolution, which mitigates
the splitting error. We have found that when we have sufficient refinement to
achieve a converged simulation, the size of the hydrodynamic timestep is already
small enough that the splitting error is minimal. For the test problem that we
will describe later, the operator splitting has no observable effect on the
physical effect we care about --- the time and location of the detonation --- at
any resolution. Consequently we do not apply any special timestep limiter based
on the rate of burning in this paper.

However, for completeness, we describe our implementation of a common approach
presented in the literature for timestep limiting based on burning, which should
be used when performing low resolution SN Ia simulations \citep{raskin:2010,hawley:2012}.
Defining the nuclear energy injection timescale $\tau_e$, and the species evolution
timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations.
This limiter is complementary to the limiter based on $f_{be}$, and can catch
certain burning regimes that limiting only based on changes in energy would miss.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. We choose to do this with a direct evaluation of the
right-hand-side of the burning network, given the current state at the time when
the timestep is evaluated. Other work in the literature has typically evaluated
this as a finite difference using the change over the last timestep. Our method
is more accurate in principle, but in practice we find no meaningful difference
between the limiting we propose and other methods of estimating the time derivatives.

\subsection{Timestep Retries}
\label{sec:retries}
